---
title: "Homework3_Final"
author: Mason Buczek
submission: 06/02/2024
git-hub link: https://github.com/Mason-Buczek/Homework3.git
format: pdf

---
## Load the necessary data, tables, and packages

```{r}
library(tidyverse)
library(readxl)
library(janitor)
library(dplyr)
library(GGally)
library(MuMIn)
library(ggeffects)
library(knitr)

# Read in the excel fike
drought_exp <- read_xlsx(path = "/Users/masonbuczek/Envs-193DS/Homework_3/Valliere_etal_EcoApps_Data.xlsx",
                         sheet = "First Harvest")
```

## Preliminary Data Cleaning

```{r}
# Cleaning the Data
drought_exp_clean <- drought_exp %>% 
  clean_names() %>% # nicer column names
  mutate(species_name = case_when( # Create new column with scientific names
    species == "ENCCAL" ~ "Encelia californica", # bush sunflower
    species == "ESCCAL" ~ "Eschscholzia californica", # California poppy
    species == "PENCEN" ~ "Penstemon centranthifolius", # Scarlet bugler
    species == "GRICAM" ~ "Grindelia camporum", # great valley gumweed
    species == "SALLEU" ~ "Salvia leucophylla", # Purple sage
    species == "STIPUL" ~ "Nasella pulchra", # Purple needlegrass
    species == "LOTSCO" ~ "Acmispon glaber" # deerweed
  )) %>% 
  relocate(species_name, .after = species) %>% # Move species_name column after species
  mutate(water_treatment = case_when( # adding column with full treatment names
    water == "WW" ~ "Well watered",
    water == "DS" ~ "Drought stressed"
  )) %>% 
  relocate(water_treatment, .after = water) # Move water_treatment column after water

# Model Construction

# Model 0: Null model
model0 <- lm(total_g ~ 1, # formula
             data = drought_exp_clean) # data frame

# Model 1: Saturated Model
model1 <- lm(total_g ~ sla + water_treatment + species_name,
             data = drought_exp_clean)
par(mfrow = c(2, 2))

# Model 2: Model total biomass as a function of SLA and water treatment
model2 <- lm(total_g ~ sla + water_treatment,
             data = drought_exp_clean)
par(mfrow = c(2, 2))

# Model 3: Model total biomass as a function of SLA and species
model3 <- lm(total_g ~ sla + species_name,
             data = drought_exp_clean)
par(mfrow = c(2, 2))

#Model 4: (My own model) Model total biomass as a function of water treatment only
model4 <- lm(total_g ~ water_treatment,
             data = drought_exp_clean)
par(mfrow = c(2, 2))

```

# Problem 1: Multiple linear regression: model selection and construction (52 points)

## Part A: Make a table or list of all the models from class and the last one you constructed on your own. Write a caption for your table. (8 points)

```{r}
# Create a data frame to store the model information
model_table <- data.frame(
  ModelNumber = 0:4,
  Predictors = c(
    "(Null Model) No predictors",
    "SLA, Water Treatment, Species",
    "SLA, Water Treatment",
    "SLA, Species",
    "Water Treatment Only")
)

# Rename the column
colnames(model_table)[1] <- "Model Number"

# Store the models in a list
models <- list(model0, model1, model2, model3, model4)

# Calculate AIC for each model
model_table$AIC <- sapply(models, AIC)

# Calculate delta AIC relative to the model with the lowest AIC
model_table$delta_AIC <- model_table$AIC - min(model_table$AIC)

# Rename the column
colnames(model_table)[4] <- "Delta AIC"

table_caption <- "Model Predictors for Predicting Total Biomass\n\n
This table provides an overview of the structure of the 5 models constructed\n
to predict total biomass. The null model does not include any predictors, while\n
the saturated model incorporates specific leaf area (SLA), water treatment, and\n
species as predictors. There are 3 additional models, two involve SLA and water\n
treatment or SLA and species as predictors with the final model having the only\n
predictor of water treatment"

# Print the table with title, caption, and centered column names
print(kable(model_table, "markdown", caption = table_caption, align = "c"))
```

## Part B: Write a 5-6 sentence “statistical methods” section. (8 points)

The central question aims to examine the influence of specific leaf area (SLA), water treatment, and species type have on specific leaf area (mm2/g), root mass (grams), and shoot mass (grams). To accomplish this goal, I constructed multiple models, including a null model and saturated model, and then proceeded to build additional models to explore different combinations of predictors. By comparing these models using criteria such as AICc and model summaries, I aimed to determine the model that best described the variation in total biomass. In selecting the final model, I focused on identifying the model with the lowest AICc value, indicating a better balance between model accuracy and model complexity. Additionally, I paid attention to the coefficients and significance levels of predictors in the final model to ensure they align with theoretical expectations and statistical significance. To visually confirm that the final model conforms to the assumptions of a linear model, To visually confirm that model1 was the best choice as it had the smallest AICc, I utilized diagnostic plots such as residual plots, Q-Q plots, and fitted vs. residuals plots to help assess if the assumptions of linearity, constant variance, and normality of residuals are met. I also examined scatter plots to assess the relationships between predictors and total biomass, ensuring they exhibit linear patterns so I can make sure I am predicting total biomass accurately. On another note, I wanted to see the effect that a singular predictor would have compared to the others that had two or three and that's why I decided to use a singular predictor in model4

## Part C: Make a visualization of the model predictions with underlying data for your “best” model. (20 points)

```{r}
# Extract model predictions
model_preds <- ggpredict(model1, 
                         terms = c("sla",
                                   "water_treatment", 
                                   "species_name"))

# Create a data frame for plotting
model_preds_for_plotting <- model_preds %>%
  rename(sla = x,
         water_treatment = group,
         species_name = facet)

# Create the visualization
ggplot() +
  # Underlying data
  geom_point(data = drought_exp_clean,
             aes(x = sla, y = total_g, color = water_treatment),
             alpha = 0.3) +
  # Model prediction 95% CI ribbon
  geom_ribbon(data = model_preds_for_plotting,
              aes(x = sla, y = predicted,
                  ymin = conf.low, ymax = conf.high,
                  fill = water_treatment),
              alpha = 0.2) +
  # Model prediction lines
  geom_line(data = model_preds_for_plotting,
            aes(x = sla, y = predicted, color = water_treatment)) +
  # Theme adjustments
  theme_classic() +
  # Remove gridlines
  theme(panel.grid = element_blank()) +
  # Facet by species
  facet_wrap(~species_name, scales = "free") +
  # Plot title
  labs(title = "Model Predictions with Underlying Data",
       x = "Specific Leaf Area (SLA)",
       y = "Total Biomass",
       color = "Water Treatment",
       fill = "Water Treatment") +
  # Legend adjustments
  scale_color_manual(values = c("Well watered" = "blue", "Drought stressed" = "red")) +
  scale_fill_manual(values = c("Well watered" = "blue", "Drought stressed" = "red")) +
  theme(legend.position = "none")
```

## Part D: Write a caption for your visualization. (6 points)

Figure 2: Total Biomass (Underlying Data) using 5 Different Model Predictor Combinations

Description:
This figure illustrates the model predictions for total biomass based on specific leaf area (SLA) and water treatment categories (Well watered vs. Drought stressed). The underlying data points represent observed total biomass values. Model predictions are depicted by solid lines, with shaded ribbons indicating the 95% confidence intervals around the predictions. The plot is faceted by species, allowing for comparisons across different plant species. Overall, the visualization provides insight into how SLA and water treatment influence total biomass in various plant species under different water conditions.

Data Citation:
Data source: Valliere, Justin; Zhang, Jacqueline; Sharifi, M.; Rundel, Philip (2019). Data from: Can we condition native plants to increase drought tolerance and improve restoration success? [Dataset]. Dryad. https://doi.org/10.5061/dryad.v0861f7

## Part E: Write a 3-4 sentence results section. (10 points)

The predictors that best described total biomass, as indicated by the selected model (Model 1), were specific leaf area (SLA), water treatment, and species. Model 1, which was identified as the best model based on statistical criteria, was a multiple linear regression model. It included specific leaf area (SLA), water treatment (Well watered vs. Drought stressed), and species as predictors. The model had a residual standard error of \( R^2 \) 0.07296 and an adjusted of 0.7223, suggesting that approximately 72.23% of the variability in total biomass was explained by the predictors in the model. Regarding differences between water treatments, the analysis revealed that plants subjected to well-watered conditions generally exhibited higher total biomass compared to those under drought-stressed conditions. This finding is supported by the coefficient estimates in Model 1, where the water treatment "Well watered" had a significantly positive coefficient estimate compared to "Drought stressed" (0.1222 vs. 0). This suggests that water treatment had a significant effect on total biomass.Furthermore, differences between species were observed in the analysis. Each species showed varying responses in total biomass to the predictors included in the model. For example, species such as Eschscholzia californica and Grindelia camporum had notably higher total biomass compared to other species, as indicated by their coefficient estimates in Model 1. In summary, the analysis suggests that specific leaf area (SLA), water treatment, and species are important predictors of total biomass in the studied plant species. Plants subjected to well-watered conditions generally exhibited higher total biomass, and different species showed varying responses to the predictors. These findings provide insights into the factors influencing total biomass and can inform strategies for managing and conserving plant populations in different environmental conditions.

# Problem 2: Affective visualization (24 points)

## Part A: Describe in words what an affective visualization could look like for your personal data (3-5 sentences). (2 points)

I really enjoyed the quilt that showed the change in temperature from a temporal sense and that was an aspect that I had not though to include in my own personal data analysis. I think it would be nice to show a image that is separated vertically that shows the change seen from a colder spring to a warmer summer. In each one of the images, it contains a drawing of me with the the more common outfit that I wore during the first half versus the latter half of my record data. I could also create a quad photo arrangement because there does not appear to be any correlation between time of year and the shoe choice I have just from looking at my data from a glance. I think this would be the most interesting visualization method to convey my data as it would show the variable that do change over time and the variables that don't. I think this way would present the observer the best interpretation of my most common outfit options seen throughout these last 6 weeks or so.

## Part B: Create a sketch (on paper) of your idea. (2 points)

While this sketch is not really indicative of the final artistic pieces that I'm am going to implement, it shows my main theme of the transition of my clothing choice from spring to summer.

## Part C: Make a draft of your visualization. (12 points)

## Part D: Write an artist statement. (8 points)

This piece shows the temporal changes over time of the combination of wearing a sweater and sweatpants versus wearing a t-shirt and shorts as spring progresses into summer. While there are no specific units on the X and Y axis, they still have meaning and representation. The "X-axis" would be time from the April 12th to May 27th and the "Y-axis" would be the probability of wearing each of the 2 specific clothing combinations. I really enjoyed the linked instagram post from "quiltinginthefog" that illustrates the global change in temperature for every year and month. I took this temporal change over time using color and applied it to my sketch. The colors of pink, green, and blue all have parallels to spring as this is when the blue spring rains are finishing up and we see the green grass with the blooming of pink flowers. This spring eventualy changes to summer and the colors of red, orange, and purple. I choose these colors and these are the major color groups that I see during the the long dusk sunsets that occur during the summer. In the middle section, we see a combination of all 6 colors which shows again the transition from spring to summer. I created this final sketch over a few days and I really got some inspiration from modern painting and the representation of colors as a greater image of representation. I used acrylic paint used a chopstick to create the mosaic pattern of dot colors and I think it came our very well especially if you can view it in person.


